Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model

### 2025-11-03 paper/paper.qmd
Pete comments

Here, we describe an simulation model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods. The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}. We compare an \acr{ABM} and a stochastic \acr{SD} model, and explore the advantages and disadvantages of each approach.

@@

$$
x_p(t) \in \{0, 1\}
$$

@@

-   $t \in \mathbb{N}_0 = \{0, 1, 2, \ldots\}$ is the time step (in days),
-   $x_p(t) = 0$ if patient $p$ is not infected at time $t$,
-   $x_p(t) = 1$ if patient $p$ is infected at time $t$.

Since the simulation's target diseases were associated with long term colonisation, and risk of onward transmission, after the initial infection event, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained infected until discharge. At each time increment, the probability of a non-infected patient acquiring infection (i.e., transitioning from a non-infected to infected state) depended on: 1) a baseline risk of *de novo* infection:

$$ 
P_\text{de novo}(t) = 1 - \text{exp}{ {( - r_0 T(t) )} },
$$

and 2) the usage of contaminated toilets:

$$
P_\text{toilet}(t) = 1 - \text{exp}{ \left( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} \right) }
$$

@@

-   $R_i(t)$ is the risk of patient $i$ acquiring infection at time $t$,

@@

-   $m$ is the increased (or decreased) relative risk of infection due to antimicrobial therapy,

-   $U_i(t)$ is the number of times patient $p$ frequents the toilet within the time interval $t$,

-   $C(t_j)$ is the contamination level of the toilet at time $t_j$, and

$$
T(t) = \begin{cases}
    m & \text{if on antimicrobial therapy} \\ % & is your "\tab"-like command (it's a tab alignment character)
    1 & \text{otherwise.}
\end{cases}
$$

The agent-based model conducted Bernoulli trials to determine if a patient became infected at each time step $t$:

$$
X_\text{de novo} \sim \text{Bernoulli}(P_\text{de novo}(t)),
$$

where $X_\text{de novo}$ = 1 if patient becomes infected *de novo* at time $t$, and 0 otherwise.

$$
X_\text{toilet} \sim \text{Bernoulli}(P_\text{toilet}(t)),
$$

where $X_\text{toilet}$ = 1 if patient becomes infected from a contaminated toilet at time $t$, and 0 otherwise.

The results of the trials were combined to determine the outcome for patient $p$ at time $t$; the patient became infected if either trial was sucessful:

$$
X_p(t) = \text{Max}(X_\text{de novo}, X_\text{toilet}),
$$

where $X_i(t)$ = 1 if patient $p$ becomes infected at time $t$, and 0 otherwise.

During the simulation level of contamination of each toilet increased when an infected patient used it, and decreased
when the toilet was cleaned.

@@

@@

To build a high-level system dynamics model, a deterministic model was firstly described in terms of the following differential equations (see @fig-sd-model), an adaptation of *SI* (susceptible--infection) models [@andersonDiscussionKermackMcKendrickEpidemic1991; @keelingModelingInfectiousDiseases2011]:

$$
\frac{dS}{dt} = (1-p)A - D \cdot \frac{S}{S + I} - \beta S - rS
$$

$$
\frac{dI}{dt} = pA - D \cdot \frac{I}{S + I} + \beta S + rS
$$

@@

-   $dS/dt$ is the rate of change of susceptible/non-infected inpatients,
-   $S$ is the number of susceptible inpatients,
-   $I$ is the number of infected inpatients,
-   $A$ is the rate of new admissions,
-   $p$ is the community infection rate,
-   $\beta$ is the transmission coefficient,
-   $r$ is the rate of random (*de novo*) infection,
-   $D$ is the rate of patient discharges.

The hospital capacity was constantly kept at 100% occupancy (i.e., patients were immediately replaced on discharge), therefore, $D = A$.

@@

To add stochasticity to the model, the number of susceptible and infected patients at each discrete time $t + 1$ was calculated by first determining the number of patients discharged:

$$
D_{\text{total}, t} \sim \text{Poisson}(D),
$$

where $D_{\text{total}, t}$ is the total number of patients (both susceptible, $D_S$, and infected $D_I$), discharged at time $t$.
Therefore, $D_{\text{total}, t} = D_{S,t} + D_{I,t}$, where:

@@

$$
A_{I,t} \sim \text{Binomial}(A_{\text{total}, t}, p),
$$

>>>>

Here, we describe an simulation model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods. The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}. We compare an \acr{ABM} and a stochastic \acr{SD} model, and explore the advantages and disadvantages of each approach for this application.

@@

$$
x_i(t) \in \{0, 1\}
$$

@@

-   $t \in \mathbb{N}_0 = \{0, 1, 2, \ldots\}$ is the time step (in days),
-   $x_i(t) = 0$ if patient $i$ is not infected at time $t$,
-   $x_i(t) = 1$ if patient $i$ is infected at time $t$.

Since the simulation's target diseases were associated with long term colonisation and prolonged risk of transmission to other patients, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained infected until discharge. At each time increment, the probability of a non-infected patient $i$ acquiring infection (i.e., transitioning from a non-infected to infected state)---$P_i(t)$---depended on: 1) a baseline risk of *de novo* infection; and 2) the usage of contaminated toilets:

$$
P_i(t) = 1 - \left[ (1 - P_{\text{de novo}, i}(t)) \cdot (1 - P_{\text{toilet}, i}(t)) \right],
$$

@@

$$ 
P_{\text{de novo}, i}(t) = 1 - \text{exp}{ {( - r_0 \cdot T_i(t) )} },
$$

$$
P_{\text{toilet}, i}(t) = 1 - \text{exp}{ \left( {-\sum_{j=0}^{U_i(t)}[C(t) \cdot T_i(t)]} \right) }
$$

-   $P_i(t)$ is the risk of patient $i$ acquiring infection at time $t$,

@@

-   $U_i(t)$ is the number of times patient $i$ frequents the toilet within the time interval $t$,

-   $C(t)$ is the contamination level of the toilet at time $t$,

-   $T(t)$ is the treatment multiplier that depends on antimicrobial therapy and is defined as:

    $$
    T(t) = \begin{cases}
      m & \text{if on antimicrobial therapy} \\
      1 & \text{otherwise, and}
    \end{cases}
    $$

-   $m$ is the increased (or decreased) relative risk of infection due to antimicrobial therapy.

In the agent-based model implementation, Bernoulli trials were conducted to determine if a patient became infected at each time step $t$:

$$
X_\text{de novo, i}(t) \sim \text{Bernoulli}(P_{\text{de novo}, i}(t)),
$$

where $X_{\text{de novo}, i}$ = 1 if patient $i$ becomes infected *de novo* at time $t$, and 0 otherwise.

Further Bernoulli trials were conducted to determine if a patient became infected from a contaminated toilet at each time step $t$:

$$
X_{\text{toilet}, i}(t) \sim \text{Bernoulli}(P_{\text{toilet}, i}(t)),
$$

where $X_{\text{toilet}, i}$ = 1 if patient $i$ becomes infected from a contaminated toilet at time $t$, and 0 otherwise.

The results of the trials were combined to determine the outcome for patient $p$ at time $t$; the patient became infected if either trial was successful:

$$
X_i(t) = \text{Max}(X_{\text{de novo}, i}, X_{\text{toilet}, i}),
$$

where $X_i(t)$ = 1 if patient $i$ becomes infected at time $t$, and 0 otherwise.

During the simulation, the contamination level $C$ of each toilet increased when an infected patient used it, and decreased when the toilet was cleaned.

@@

Each patient's admission duration was sampled from a Poisson distribution with a configurable mean (in days). Upon discharge, patients were removed from the simulation, and replaced with new admissions to maintain 100% occupancy. New admissions had a probability of being infected based on the community infection rate parameter.

@@

To build a high-level stochastic system dynamics model, the model was first described in terms of the following differential equations (see @fig-sd-model), an adaptation of *SI* (susceptible--infection) models [@andersonDiscussionKermackMcKendrickEpidemic1991; @keelingModelingInfectiousDiseases2011]:

$$
\frac{dS}{dt} = (1-\lambda)A - D \cdot \frac{S}{S + I} - S (\beta + r)
$$

$$
\frac{dI}{dt} = \lambda A - D \cdot \frac{I}{S + I} + S (\beta + r)
$$

@@

-   $S$ is the number of susceptible inpatients,
-   $I$ is the number of infected inpatients,
-   $dS/dt$ and $dI/dt$ are the rates of change of susceptible/non-infected and infected inpatients, respectively,
-   $A$ is the number of new admissions,
-   $D$ is the number of patient discharges,
-   $\lambda$ is the community infection rate,
-   $\beta$ is the transmission coefficient,
-   $r$ is the rate of random (*de novo*) infection.

The hospital capacity was constantly kept at 100% occupancy (i.e., patients were immediately replaced on discharge), therefore, $d = a$.

@@

Next, the number of susceptible and infected patients at each discrete time $t + 1$ was calculated by first determining the number of patients discharged:

$$
D_{\text{total}, t} \sim \text{Poisson}(d),
$$

where $d$ is the discharge rate and $D_{\text{total}, t}$ is the total number of patients (both susceptible, $D_S$, and infected $D_I$), discharged at time $t$.

Therefore, $D_{\text{total}, t} = D_{S,t} + D_{I,t}$, where:

@@

$$
A_{I,t} \sim \text{Binomial}(A_{\text{total}, t}, \lambda),
$$
